---
title: "Assignment 3 workflow"
author/collab: Heather Childers & Sofia Ingersoll
output: html_document
date: "2023-11-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Exploring the Socioeconomic Impacts of the 2021 Blackout in Houston, Texas

## Overview

In February 2021, the state of Texas suffered a major power crisis, which came about as a result of three severe winter storms sweeping across the United States on February 10--11, 13--17, and 15--20."[^1] For more background, check out these [engineering](https://www.youtube.com/watch?v=08mwXICY4JM&ab_channel=PracticalEngineering) and [political](https://www.youtube.com/watch?v=Zcrsgdl_hP0&ab_channel=Vox) perspectives.

[^1]: Wikipedia. 2021. "2021 Texas power crisis." Last modified October 2, 2021. <https://en.wikipedia.org/wiki/2021_Texas_power_crisis>. - `gis_osm_roads_free_1.gpkg`

The goal of this repository is to estimate the number of homes in Houston that lost power as a result of the first two storms. We will also investigate if socioeconomic factors are predictors of communities recovery from a power outage.

This analysis will be based on remotely-sensed night lights data, acquired from the [Visible Infrared Imaging Radiometer Suite (VIIRS)](https://en.wikipedia.org/wiki/Visible_Infrared_Imaging_Radiometer_Suite) onboard the Suomi satellite. In particular, we will use the VNP46A1 to detect differences in night lights before and after the storm to identify areas that lost electric power.

To determine the number of homes that lost power, you link (spatially join) these areas with [OpenStreetMap](https://www.openstreetmap.org/#map=4/38.01/-95.84) data on buildings and roads. To determine the number of homes that lost power, you link (spatially join) these areas with [OpenStreetMap](https://www.openstreetmap.org/#map=4/38.01/-95.84) data on buildings and roads.

To investigate potential socioeconomic factors that influenced recovery, we will link this analysis with data from the US Census Bureau. To investigate potential socioeconomic factors that influenced recovery, we will link this analysis with data from the US Census Bureau.

#### Skills used in this Repo:

-   load vector/raster data

-   simple raster operations

-   simple vector operations

-   spatial joins

### Data

#### Night lights

Use NASA's Worldview to explore the data around the day of the storm. There are several days with too much cloud cover to be useful, but 2021-02-07 and 2021-02-16 provide two clear, contrasting images to visualize the extent of the power outage in Texas.

VIIRS data is distributed through NASA's [Level-1 and Atmospheric Archive & Distribution System Distributed Active Archive Center (LAADS DAAC)](https://ladsweb.modaps.eosdis.nasa.gov/). Many NASA Earth data products are distributed in 10x10 degree tiles in sinusoidal equal-area projection. Tiles are identified by their horizontal and vertical position in the grid. Houston lies on the border of tiles h08v05 and h08v06. We therefore need to download two tiles per date.

As you're learning in EDS 220, accessing, downloading, and preparing remote sensing data is a skill in it's own right! To prevent this assignment from being a large data wrangling challenge, we have downloaded and prepped the following files for you to work with, stored in the `VNP46A1` folder.\

-   `VNP46A1.A2021038.h08v05.001.2021039064328.h5.tif`: tile h08v05, collected on 2021-02-07\
-   `VNP46A1.A2021038.h08v06.001.2021039064329.h5.tif`: tile h08v06, collected on 2021-02-07\
-   `VNP46A1.A2021047.h08v05.001.2021048091106.h5.tif`: tile h08v05, collected on 2021-02-16\
-   `VNP46A1.A2021047.h08v06.001.2021048091105.h5.tif`: tile h08v06, collected on 2021-02-16\
-   `VNP46A1.A2021038.h08v05.001.2021039064328.h5.tif`: tile h08v05, collected on 2021-02-07\
-   `VNP46A1.A2021038.h08v06.001.2021039064329.h5.tif`: tile h08v06, collected on 2021-02-07\
-   `VNP46A1.A2021047.h08v05.001.2021048091106.h5.tif`: tile h08v05, collected on 2021-02-16\
-   `VNP46A1.A2021047.h08v06.001.2021048091105.h5.tif`: tile h08v06, collected on 2021-02-16

#### Roads

Typically highways account for a large portion of the night lights observable from space (see Google's [Earth at Night](https://earth.google.com/web/@27.44405464,-84.7693044,206.63660162a,8916361.52264659d,35y,0h,0t,0r/data=CiQSIhIgMGY3ZTJkYzdlOGExMTFlNjk5MGQ2ZjgxOGQ2OWE2ZTc)). To minimize falsely identifying areas with reduced traffic as areas without power, we will ignore areas near highways.

[OpenStreetMap (OSM)](https://planet.openstreetmap.org/) is a collaborative project which creates publicly available geographic data of the world. Ingesting this data into a database where it can be subsetted and processed is a large undertaking. Fortunately, third party companies redistribute OSM data. We used [Geofabrik's download sites](https://download.geofabrik.de/) to retrieve a shapefile of all highways in Texas and prepared a Geopackage (`.gpkg` file) containing just the subset of roads that intersect the Houston metropolitan area.  Typically highways account for a large portion of the night lights observable from space (see Google's [Earth at Night](https://earth.google.com/web/@27.44405464,-84.7693044,206.63660162a,8916361.52264659d,35y,0h,0t,0r/data=CiQSIhIgMGY3ZTJkYzdlOGExMTFlNjk5MGQ2ZjgxOGQ2OWE2ZTc)). To minimize falsely identifying areas with reduced traffic as areas without power, we will ignore areas near highways.

-   `gis_osm_roads_free_1.gpkg` [OpenStreetMap (OSM)](https://planet.openstreetmap.org/) is a collaborative project which creates publicly available geographic data of the world. Ingesting this data into a database where it can be subsetted and processed is a large undertaking. Fortunately, third party companies redistribute OSM data. We used [Geofabrik's download sites](https://download.geofabrik.de/) to retrieve a shapefile of all highways in Texas and prepared a Geopackage (`.gpkg` file) containing just the subset of roads that intersect the Houston metropolitan area.

#### Houses

We can also obtain building data from OpenStreetMap. We again downloaded from Geofabrick and prepared a GeoPackage containing only houses in the Houston metropolitan area.\

-   `gis_osm_buildings_a_free_1.gpkg`
-   `gis_osm_buildings_a_free_1.gpkg`

#### Socioeconomic

We cannot readily get socioeconomic information for every home, so instead we obtained data from the [U.S. Census Bureau's American Community Survey](https://www.census.gov/programs-surveys/acs) for census tracts in 2019. The *folder* `ACS_2019_5YR_TRACT_48.gdb` is an ArcGIS ["file geodatabase"](https://desktop.arcgis.com/en/arcmap/latest/manage-data/administer-file-gdbs/file-geodatabases.htm), a multi-file proprietary format that's roughly analogous to a GeoPackage file.\

You can use `st_layers()` to explore the contents of the geodatabase. Each layer contains a subset of the fields documents in the [ACS metadata](https://www2.census.gov/geo/docs/maps-data/data/tiger/prejoined/ACSMetadata2011.txt).\

The geodatabase contains a layer holding the geometry information, separate from the layers holding the ACS attributes. You have to combine the geometry with the attributes to get a feature layer that `sf` can use. The geodatabase contains a layer holding the geometry information, separate from the layers holding the ACS attributes. You have to combine the geometry with the attributes to get a feature layer that `sf` can use.

### Import Relevant Libraries

```{r libraries for data analysis}
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
library(geodata)
library(stars)
library(sf)
library(tidyverse)
```

## Read in the Data

Read in each of the datasets described above.

#### Nightlights Data:

Use the `read_stars()` function to read in the data as a stars object

```{r pressure, echo=FALSE, }
# Read in the Nightlights Data
tile5_Feb7 <- read_stars("data/VNP46A1.A2021038.h08v05.001.2021039064328.tif") 
tile6_Feb7 <- read_stars("data/VNP46A1.A2021038.h08v06.001.2021039064329.tif") 
tile5_Feb16 <- read_stars("data/VNP46A1.A2021047.h08v05.001.2021048091106.tif") 
tile6_Feb16 <- read_stars("data/VNP46A1.A2021047.h08v06.001.2021048091105.tif") 

```

Check the class of each object to ensure the data was read in as a stars object

```{r}
class(tile5_Feb7)
class(tile5_Feb16)
class(tile6_Feb7)
class(tile6_Feb16)
```

#### Highways Data:

We'll be using queries to to access data from available databases. SQL is a Standardized Query Language that allows users on different operating systems and programming languages to access relational databases. The queries are the commands that access the data.

Use the below query to access the Highways data

```{r}
#Use a query to import highways data
highway_query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"
#Use st_read to read in the highway data as a spatial object
highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = highway_query, quiet = TRUE)
```

#### Housing Data:

Use the below query to access the Housing data:

```{r}
#Use a query to import housing data
building_query <- "SELECT* 
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')"
#Use st_read to read in the housing data as a spatial object
buildings <-  st_read("data/gis_osm_buildings_a_free_1.gpkg", query = building_query, quiet = TRUE)
```

#### Socioeconomic Data:

use the `st_read()` function to read in the Socioeconimic data

```{r}
#Use st_read to read in the geometry layer of the socioeconomic data
geometry <- st_read('data/ACS_2019_5YR_TRACT_48_TEXAS.gdb/', layer = "ACS_2019_5YR_TRACT_48_TEXAS")
#Use st_read to read in the socioeconomic data
socioeconomic <- st_read('data/ACS_2019_5YR_TRACT_48_TEXAS.gdb/', layer = 'X19_INCOME')
```

## Initial Data Exploration and Manipulation

#### Nightlights Data:

Combine tiles from the same day to create 2 comparable images

```{r}
#Use st_mosaic to combine the two stars objects
Feb_7 = st_mosaic(tile5_Feb7, tile6_Feb7)
Feb_16 = st_mosaic(tile5_Feb16, tile6_Feb16)
```

checks:

```{r}
#Check that the data was combined properly by plotting the image
plot(Feb_7)
#Re-check the class of the new combined object
class(Feb_7)

#Check that the data was combined properly by plotting the image
plot(Feb_16)
#Re-check the class of the new combined object
class(Feb_16)

```

#### Highway Data:

In the following section we will re-project the data to have a consistent CRS and add a 200meter buffer around each highway. NOTE: toggling between the original highway data and the highway with buffer data allows you to see the line thickness change
#this ensures we actually made the buffer and added it to the geometry successfully.

```{r}
#Check the CRS of the highways data
crs(highways)
#Transform the highways data to the EPSG 3083 projection
highways <- highways %>% 
  st_transform(crs = 'EPSG:3083')
#Plot the initial data
plot(highways$geom)
#Add a 200m buffer around the highways so light coming from highways doesn't distort the data
buffer <- highways %>% 
  st_buffer(dist = 200)
#Use st_union to join the highways data and the buffer into one cohesive geometry
highways_with_buffer <- buffer %>% 
  st_union()
#Plot the highways data with the buffer to see the new data
plot(highways_with_buffer)
```

#### Socioeconomic Data:

Select only the GEOID and Median Income columns from the imported socioeconomic data and rename those columns for ease of use in the future.

```{r}
#Select and rename the data
socioeconomic <- socioeconomic %>% 
  dplyr::select(GEOID, B19013e1) %>% 
  rename(GEOID_Data = GEOID,
    median_income = B19013e1) 

```

Join the downloaded Socioeconomic data to the relevant geometries:

```{r}
#Check the CRS of the original Geometry
crs(geometry)
#Join the geometery layer to the Census tract and Median Income data
#Reproject the data to the EPSG 3083 projection
census_income <- left_join(geometry, socioeconomic) %>% 
  st_transform('EPSG:3083')

#Check that the data joined correctly by viewing the head of the data
head(census_income)
```

## Task 1: Finding the locations of Blackouts

We'll start by creating a blackout mask:

```{r}
#find the change in night lights intensity (presumably) caused by the storm
#assume that any location that experienced a drop of more than 200 nW cm-2sr-1 experienced a blackout
mask <- (Feb_7 - Feb_16) > 200

#assign NA to all locations that experienced a drop of less than 200 nW cm-2sr-1
mask[mask == FALSE] = NA
```

Then vectorize the mask:

```{r}
#use st_as_sf() to vectorize the blackout mask
blackout <- st_as_sf(mask)
#fix any invalid geometries using st_make_valid
blackout<- st_make_valid(blackout)
```

Finally, crop the vectorized map to our region of interest:

NOTE: We will assume the boundary of Houston can be defined by the following coordinates:

-   (-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29)

```{r}
#turn these coordinates into a polygon using st_polygon
Houston <- st_polygon(list(rbind(c(-96.5, 29),c(-96.5, 30.5),c(-94.5, 30.5), c(-94.5, 29),c(-96.5, 29))))
#convert the polygon into a simple feature collection using st_sfc() and assign a CRS
Houston <- st_sfc(Houston) %>% 
  st_set_crs(4326)
```

```{r}
#crop (spatially subset) the blackout mask to our region of interest
cropped_mask<- st_crop(blackout,Houston)
#Use the plot function to see that your data has been subsetted
ggplot()+
  geom_sf(data = cropped_mask$geometry)
#re-project the cropped blackout dataset to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area)
cropped_mask <- cropped_mask %>% 
  st_transform('EPSG:3083') %>% 
  st_make_valid()


```

To avoid mistaking differences in traffic for a blackout, remove the highways from the mask:

```{r}
#Use st_difference to select the areas that are within the cropped mask and outside the buffer
cropped_mask_no_highways <- st_difference(cropped_mask,highways_with_buffer)
#check to see that the subset worked by comparing the number of observations in each dataset:
print(nrow(cropped_mask))
print(nrow(cropped_mask_no_highways))
```

Note: there are 7256 observations in our original dataset, and there are only 7247 observations in the new dataset.

## Find homes impacted by blackouts

```{r}
#Reproject the buildings data to be the same CRS
buildings <- buildings %>% 
  st_transform('EPSG:3083')
#Use st_within to select buildings that are within our blackout mask
homes_with_blackouts <- st_intersection(buildings,cropped_mask_no_highways)
```

```{r}
#check the new data by plotting the homes with blackouts
plot(homes_with_blackouts$geom)
```

```{r}
#Find how many homes were affected
nrow(homes_with_blackouts)
```

There were 168,874 homes affected by the blackouts in Houston.

## Investigate socioeconomic factors

Find which census tracts had blackouts:

```{r}
#Check the CRS of the census_income data
crs(census_income)
#Check the CRS of the houston polygon
crs(Houston)
#Transform the Houston Polygon to the same CRS as the census income info
Houston <- Houston %>% 
  st_transform('EPSG:3083')

# Use st_crop to crop the Texas Census Data to Houston Boundaries
census_houston <- st_crop(census_income, Houston)

#Plot the data to check it's cropped
plot(census_houston$Shape)
```

```{r}
#Combine the Census data with the blackout data, set available values equal to ture
combined <- st_join(census_houston, homes_with_blackouts, left = TRUE)
#Set the NA Values equal to False
combined$VNP46A1.A2021038.h08v05.001.2021039064328.tif[is.na(combined$VNP46A1.A2021038.h08v05.001.2021039064328.tif)] <- "FALSE" 



```

```{r}
#Create a new variable to identify just the tracts that experienced blackouts
trac_with_blackout <- combined %>% 
  filter(combined$VNP46A1.A2021038.h08v05.001.2021039064328.tif == TRUE)
##Create a new variable to identify just the tracts that didnt experience blackouts
trac_without_blackout <- combined %>% 
  filter(combined$VNP46A1.A2021038.h08v05.001.2021039064328.tif == FALSE)
```

-   create a map of median income by census tract, designating which tracts had blackouts

```{r}
#Create a map of median income by census tract, designating which tracts had blackouts
tm_shape(census_houston) + 
  tm_polygons("median_income", 
          title = "Median Income ($)",
          palette = "magma") +
  tm_shape(trac_with_blackout) +
  tm_dots() 

```

```{r}
  tm_shape(trac_with_blackout) +
  tm_polygons("median_income", 
          title = "Median Income ($)",
          palette = "RdPu") 
```

```{r}
  tm_shape(trac_without_blackout) +
  tm_polygons("median_income", 
          title = "Median Income ($)",
          palette = "RdPu") 
```

-   plot the distribution of income in impacted and unimpacted tracts

```{r}
ggplot(combined, aes(x=VNP46A1.A2021038.h08v05.001.2021039064328.tif, y=median_income))+
  geom_boxplot()+
  labs(x = "Experienced a Blackout",
       y = 'Median Income ($)',
       title = "Median Income spearated by Blackout Status")

```

This distribution was originally surprising to me because I expected there to be some information about lower-income census tracts being affected by the blackouts. I noticed from this data that so many buildings were affected by the blackouts; they weren't based on income. The income from the census tracts that didn't experience a blackout was lower than the census tracts that did because so few census tracts avoided the blackouts. The main limitation that I could identify in this study is that the original data wasn't available in the same projection. In order to join the spatial data, we had to reproject the data, which required the use of the st_make_valid() function in a few places in order for the code to run. Ideally, we could find data that was available in the same projection, so the geometries were more consistent.
